---
title: "My Beginners Analysis"
author: "Denis Motruk"
date: "January 17, 2020"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

To get a better understanding and feel for R and applied statistics as a whole I decided to create my own data analysis project.
The Data I will be using can be found here https://www.kaggle.com/russellyates88/suicide-rates-overview-1985-to-2016

```{r include=FALSE}
library(readxl)
library(data.table)
library(ggplot2)
library(tidyverse)
library(gvlma)
library(car)

```
##Importing the Data and cleaning 

Had to get rid of empty HDI values and their associated rows
```{r}
suicides <- fread("suicide data.csv")
hdi <- suicides[`HDI for year` != 'NA']

all <- hdi[,lapply(.SD, mean), by=.(country, year, sex),.SDcols = c("suicides/100k pop", "HDI for year", "gdp_per_capita ($)")]
no_null <- all[`suicides/100k pop` != '0']

no_null2 <- no_null %>% 
  mutate(female = ifelse(grepl("female", sex), "1", "0"))

```

##Modelling Suicide Rates
I need to be careful however, since there exists a relationships between HDI and GDP per capita.
```{r}
rate <- no_null2$`suicides/100k pop`
hdi <- no_null2$`HDI for year`
gdp <- no_null2$`gdp_per_capita ($)`
sex <- no_null2$female

plot(hdi, gdp)

```
To deal with this I split the model to see individual effects

##Rate modelled by HDI 
```{r}
mdl <- lm(log(rate)~hdi)
summary(mdl)

plot(hdi, log(rate))
abline(mdl, col='red', pch=19)
```
We can see that there is a slight relationship with HDI and Suicide rates in the world

##Rate modelled by GDP 
```{r}
mdl2 <- lm(log(rate)~gdp)
summary(mdl2)
plot(gdp, log(rate))
abline(mdl2, col='red', pch=19)
```
As we can see GDP per capita also has a relationship with Suicide rates


##Compare means of the different year chunks

I had to apply a log transformation to better satisfy the assumptions of a linear model

```{r}
yrs1 <- subset(no_null2, format(as.numeric(year))<=2000)
yrs2 <- subset(no_null2, format(as.numeric(year))>2000)


y1rate <- yrs1$`suicides/100k pop`
y1hdi <- yrs1$`HDI for year`


yr1model <- lm(log(y1rate) ~ (y1hdi))
summary(yr1model)



```


```{r}
y2rate <- yrs2$`suicides/100k pop`
y2hdi <- yrs2$`HDI for year`

yr2model <- lm(log(y2rate) ~ (y2hdi))
summary(yr2model)
```


```{r}
boxplot(y1rate, y2rate)
t.test(y1rate, y2rate)
```


##Conclusions
Both GDP per capita and HDI are correlated with a higher suicide rate for a given country. However neither are great predictors of suicde rates. Men have higher suicide rates than women. Additionally we can see that suicide rates (globally) have been going down.

Barring me having done something terribly wrong, I believe it makes sense that neither of these varibales happen to predict suicide rates well. As we often see, there are many factors that going into such a decision. However, these findings help us identify populations of people who may be at higher risk and thus preventitive measures can be taken.

#Next Steps
Investigate how HDI effects suicide rates in wealthy vs poor countries

#Note
This is by no means a comprehensive study. Out of all my research I found this to provide the most insight into the data so far. I plan to expand this page and its findings as I learn more.
